\documentclass[10pt]{article}
\usepackage{amsmath}
\usepackage{setspace}
\usepackage{hyperref}


\setlength{\textheight}{9in} \setlength{\topmargin}{-.5in}
\setlength{\textwidth}{6.5in} \setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}

\title{COMP 161 - Main Sources} 
\author{ James \textit{Logan} Mayfield }
\date{Spring 2015}

\begin{document}
\maketitle

\begin{enumerate}
\item \textbf{Linux CLI (BASH)}
\begin{itemize}
\item Primary 

Shotts, William. \textit{The Linux Command Line}. No Starch Press. 2012. 
\newline \url{http://linuxcommand.org/tlcl.php}

\item Additional Sources

Shaw, Zed. \textit{The Command Line Crash Course: Controlling Your Computer From The Terminal}. Dec 2011. 
\newline \url{http://cli.learncodethehardway.org/book/}
\vspace{.25in}

Nodwell, Eric. \textit{Introduction to Commandline Linux}. 2003.
\newline \url{http://www.phas.ubc.ca/~mbelab/computer/linux-intro/html/}
\end{itemize}

\item \textbf{EMACS}
\begin{itemize}
\item Primary

Start EMACS. Enter the command \textit{Ctrl-h t} to begin the interactive tutorial.
\newline

Emacs Reference Sheet. \newline
\url{http://www.damtp.cam.ac.uk/user/sje30/ess11/resources/emacs-refcard.pdf}	


\item Additional Sources

Wacelna, Keith. \textit{A Tutorial Introduction to GNU Emacs}. 2009.
\newline \url{http://www2.lib.uchicago.edu/keith/tcl-course/emacs-tutorial.html}
\vspace{.25in}

Pitts, Robert I. \textit{Using Emacs for Programming}. 2000. 
\newline \url{http://www.cs.bu.edu/teaching/tool/emacs/programming/} 
\vspace{.25in}

\end{itemize}

\item \textbf{C++}

\textit{CPlusPlus.com} 
\newline \url{http://cplusplus.com}
\vspace{.25in}

Stroustrup, Bjarne. \textit{A Tour of C++}. Addison-Wesley. New Jersey. 2014.  

\item \textbf{GCC/G++}

Chen, Bee-Chunk. \textit{Compiling C and C++ Programs}.
\newline \url{http://pages.cs.wisc.edu/~beechung/ref/gcc-intro.html}
\vspace{.25in}

GNU GCC Manual \newline
\url{http://gcc.gnu.org/onlinedocs/gcc-4.8.2/gcc/}

\item \textbf{make}

GNU Make Manual \newline
\url{http://www.gnu.org/software/make/manual/make.html}

\item \textbf{GDB}

GDB Manaul \newline
\url{http://sourceware.org/gdb/current/onlinedocs/gdb/}
\vspace{.25in}

Hall, Brian ``Beej Jorgensen''. \textit{Beej's Quick Guide to GDB} \newline
\url{http://beej.us/guide/bggdb/}
\vspace{.25in}

Lancaster, Glenn. \textit{Quick Guide to GDB} \newline
\url{http://condor.depaul.edu/glancast/373class/docs/gdb.html}
\vspace{.25in}

\item \textbf{Valgrind (memcheck,callgrind,cachegrind)}

Official Quick Start Guide \newline
\url{http://valgrind.org/docs/manual/QuickStart.html}
\vspace{.25in}

Official Manual \newline
\url{http://valgrind.org/docs/manual/manual.html}

\item \textbf{gcov}

Official Manual \newline
\url{http://gcc.gnu.org/onlinedocs/gcc/Gcov.html}

\item \textbf{Git}

Chacon, Scott. \textit{Pro Git}. \url{http://git-scm.com/book}


\end{enumerate}

\end{document}